Nonlinear Tidal Dissipation in Binary White Dwarfs

### 2019-02-19 proposal/FINESST/Su_FINESST_2019.tex
typo fixes

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking process exhibits new behavior that
cannot be described by linear and weakly nonlinear
theory\cite{winters1994,barker_ogilvie}. Such fully nonlinear numerical
simulations have not been performed for WDs.

@@

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales that causes wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear Internal Gravity Wave (IGW) breaking
in binary WDs.} There are three specific goals of my proposed research:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent of wave breaking will furnish a simple
        yet effective parameterization of tidal dissipation in a range of WD
        models. \autoref{s:2} describes the steps required to acheive this goal.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will study the impact of tidal heating
        on the luminosity of WDs in binaries and explore the possibility of
        producing of observable flares. \autoref{s:3} details my plan.

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: tidal forces from the companion excites IGWs in the deep
envelope of the WDs. These IGW propagate outwards and undergo wave breaking in
the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the binary orbit. A similar process also
operates in binaries consisting of early type stars\cite{zahn75,gn89},
the only major difference being in the specifics of wave
excitation\footnote{While IGWs in massive stars are excited at their
radiative-convective boundaries, the excitation of IGWs in WDs is more gradual
and is associated with sharp composition changes in the stellar
envelope\cite{fullerII}.}. Nevertheless, direct numerical simulation of the wave
breaking process has not been performed in either of these systems. Since wave
breaking is a strongly nonlinear phenomenon, where a larger wave breaks down
into many smaller-scale waves, numerical simulation is paramount to an accurate
understanding of the tidal dissipation process.

IGW breaking has been studied in atmospheric sciences. The wave
breaking process proceeds as follows: Initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}; after the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer forms. Analytical
calculations show that the IGW is nearly completely absorbed at this critical
layer in the linear approximation and endows the atmosphere with a mean
horizontal flow\cite{booker_bretherton,hazel}. However, when this mean flow
absorption was numerically studied including full nonlinear interactions, new
phenomena not described by the linear theory (reflection off the critical layer
and sharpening of the mean flow) were observed\cite{jones_num,winters1994}; this
nonlinear behavior significantly affectedthe evolution of the atmosphere over
time. \emph{This highlights the importance of numerical simulation in capturing
the wave breaking process.}

@@

It is clear that full numerical modeling of IGW breaking is necessary to fully
characterize tidal dissipation. The first aim of my proposal is to
\textbf{extend my preliminary results concerning IGW breaking in stratified
atmospheres to characterize the dynamics of nonlinear IGW breaking in realistic
WD models.} Via numerical simulation, I will develop models for tidal
dissipation that can be used to study long-term WD evolution under tidal heating
without expensive hydrodynamical simulations. I will continue my work in the
following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        spherical geometries to capture tidal effects in WD binaries. I will
        continue to use Dedalus, which supports spherical coordinates. Although
        3D simulations are necessarily more complex than my 3D simulations of
        stratified atmospheres, the underlying dynamics of IGW breaking are the
        same.

>>>>

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory\cite{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

Characterizing the location and spatial extent of tidal dissipation in WD
binaries will require numerical simulation to capture the turbulent cascade to
small scales that causes wave breaking. \textbf{I propose to study the dynamical
effects of tidal dissipation via nonlinear internal gravity wave (IGW) breaking
in binary WDs.} There are three specific goals of my proposed research:
\begin{itemize}
    \item Characterize the location, spatial extent, and other properties of
        wave breaking in realistic WD models via direct numerical simulation.
        The location and spatial extent of wave breaking will furnish a simple
        yet effective parameterization of tidal dissipation in a range of WD
        models. \autoref{s:2} describes the steps required to acheive this goal.

    \item Predict signatures of tidal dissipation over a wide range of possible
        WD systems. In particular, I will study the impact of tidal heating
        on the luminosity of WDs in binaries and explore the possibility of
        producing of observable flares. \autoref{s:3} details my plan to predict
        observational manifestations of tidal heating.

@@

The current understanding of tidal synchronization in WD binaries is laid out
in~\cite{fullerII}: Tidal forces from the companion excites IGWs in the deep
envelope of the WDs. These IGW propagate outwards and undergo wave breaking in
the outer envelope of the WD, locally depositing angular momentum and
synchronizing the WD spin to the binary orbit. A similar process also
operates in binaries consisting of early type stars\cite{zahn75,gn89},
the only major difference being in the specifics of wave
excitation\footnote{While IGWs in massive stars are excited at their
radiative-convective boundaries, the excitation of IGWs in WDs is more gradual
and is associated with sharp composition changes in the stellar
envelope\cite{fullerII}.}. Nevertheless, direct numerical simulation of the wave
breaking process has not been performed in either of these systems. Since wave
breaking is a strongly nonlinear phenomenon, where a larger wave breaks down
into many smaller-scale waves, numerical simulation is paramount to an accurate
understanding of the tidal dissipation process.

IGW breaking has been studied in atmospheric sciences. The wave
breaking process proceeds as follows: Initially, as the IGW reaches nonlinear
amplitudes, it breaks down via the parametric subharmonic instability and
transfers energy and angular momentum from the wave to the mean flow of the
fluid\cite{drazin}. After the mean flow velocity reaches the
horizontal phase velocity of the IGW, a critical layer forms. Analytical
calculations show that the IGW is nearly completely absorbed at this critical
layer in the linear approximation and endows the atmosphere with a mean
horizontal flow\cite{booker_bretherton,hazel}. However, when this mean flow
absorption was numerically studied including full nonlinear interactions, new
phenomena not described by the linear theory (reflection off the critical layer
and sharpening of the mean flow) were observed\cite{jones_num,winters1994}; this
nonlinear behavior significantly affected the evolution of the atmosphere over
time. \emph{This highlights the importance of numerical simulation in capturing
the wave breaking process.}

@@

It is clear that full numerical modeling of IGW breaking is necessary to fully
capture tidal dissipation. The first aim of my proposal is to \textbf{extend my
preliminary results concerning IGW breaking in stratified atmospheres to
characterize the dynamics of nonlinear IGW breaking in realistic WD models.} Via
numerical simulation, I will develop models for tidal dissipation that can be
used to study long-term WD evolution under tidal heating without expensive
hydrodynamical simulations. I will continue my work in the following stages:
\begin{itemize}
    \item I will perform simulations examining the validity of my results in
        spherical geometries to capture tidal effects in WD binaries. I will
        continue to use Dedalus, which supports spherical coordinates. Although
        3D simulations are necessarily more complex than my 3D simulations of
        stratified atmospheres, the underlying dynamics of IGW breaking are the
        same.
